---
title: "Workshop 8 TEMPLATE"
subtitle: "9 AM"
author: "Taylor Platz"
date: "2026-02-27"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---
[link](https://taylorplatz.github.io/workshop-08_winter-2026/code/workshop-code.html)

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 6 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschscholzia californica_   | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here

drought_exp <- read_xlsx(
  path = here(
    "data",
    "Valliere_etal_EcoApps_Data.xlsx"
  ),
  sheet = "First Harvest"
)

```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species, saving all colors as objects
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# creating a new object called "drought_exp_clean"
drought_exp_clean <- drought_exp |> 
  
  # cleaning up column names
  clean_names() |>  
  # adding a new column called species name and changing the species abbreviation to the full names
  mutate(species_name = case_match(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |>  
  # add new column for treatment and use the abbreviations and changing them to the corresponding names
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  ))  |> 
  # making sure that categorical variables are factors, and setting factor level order, ordered in smallest to biggest mass and telling R that these names are important in a specif order
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  #
  select(species_name, water_treatment, sla, total_g) 
```

Insert code to double check the structure of the data frame here.  
```{r}
#| label: checking structure
str(drought_exp_clean)

```


```{r}
   
slice_sample(
  drought_exp_clean,
  n = 10 
)
```

Insert code to display 10 random rows from the data frame here.  

# 3. Visualizing data

## What is the relationship between SLA and plant mass?  


```{r}
# base layer: ggplot 
ggplot(data = drought_exp_clean,
  aes(x = sla,
      y = total_g)) + 
  geom_point(
    color = "forestgreen"
  ) +
  labs(
    y = "Total Mass (g)",
    x = "Specific Leaf Area (cm\U00B2/g)",
    title = "As SLA increases, total mass increases")
  
```


```{r}

```

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 





## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.

```{r}
ggplot(data = drought_exp_clean,
       aes(
         x = water_treatment,
         y = total_g,
         color = water_treatment)) +
  geom_jitter(
    height = 0,
    width = 0.2
  ) +
  scale_color_manual(values = c(
    "Drought Stressed" = ds_col,
    "Well Watered" = ww_col
  )) +
  labs(x = "Water treatment",
       y = "Total mass (g)",
       title = "On average, plant mass in well watered tretment \n is higher that plant mass in drought stressed \n treatment")
  
  

```






## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass _and_ wrap the long species names using the `label_wrap()` function from the `scales` package.
```{r}
ggplot(data = drought_exp_clean,
       aes(x = species_name,
           y = total_g,
           color = species_name)) +
  geom_jitter(
    height = 0,
    width = 0.2) +
  scale_color_manual(values = c(
    "Lotus scoparius" = lotsco_col,
    "Penstemon centanthifolius" = pencen_col
  ))
```








# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model
model0 <- lm(
  total_g ~ 1,
  data = drought_exp_clean
)

# model 1: all predictors
model1 <- lm(
  total_g ~ sla + water_treatment + species_name,
  data = drought_exp_clean
)

# model 2: SLA and water treatment
model2 <- lm(
  total_g ~ sla + water_treatment,
  data = drought_exp_clean
)

# model 3: SLA and species
model3 <- lm(
  total_g ~ sla + species_name,
  data = drought_exp_clean
)

# model 4: water treatment and species
model4 <- lm(
  total_g ~ water_treatment + species_name,
  data = drought_exp_clean
)

# model 5: SLA
model5 <- lm(
  total_g ~ sla,
  data = drought_exp_clean
)

# model 6: water treatment
model6 <- lm(
  total_g ~ water_treatment,
  data = drought_exp_clean
)


# model 7: species
model7 <- lm(
  total_g ~ species_name,
  data = drought_exp_clean
)

```

## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
AICc(
  model0,
  model1,
  model2,
  model3,
  model4,
  model5,
  model6,
  model7
) |> 
  arrange(AICc)
```

the best model to describe variation in plant mass only includes water treatment and species

## Model diagnostics

Insert code to look at the diagnostic plots for the best model.


```{r}
#| label: checking diagnostics

par(mfrow = c(2,2))
plot(model4)
```

Diagnostic plots look good
Risiduals (standing in for error) look normally distributed based on QQ plot
Residuals look homoscedastic

no outliers 

## Summary

```{r}
#| label: looking at model summary
summary(model4)
```

Insert code to display the summary for the best model. 

- two categorical predictors
- each estimate is comparing the 

**What are the reference levels?**  

Drought stressed is the reference level for water treatment
Lotus scoparius for species (in which plant mass is smallest)

**What does the `(Intercept)` represent?**  

model estimate when all predictors = 0, but because they are catagorical, its is the model prediction for plant mass for lotus scoparius in the drought stressed treatment. The model predicts that Lotus Scoparius mass int he drought stressed treatment will be 0.055 $\pm$ 0.025 (SE) g. 

**What does `water_treatmentWell watered` represent?**

This is the model prediction for the difference in the difference in Lotus mas between the well watered and drought stressed treatments. The model predicts that this difference is 0.12 $\pm$ 0.017 g. 


**What does `species_namePenstemon centranthifolius` represent?**  
This is the model prediction for the difference between Panstemon mass in the well watered treatment and Lotus mass in the well watered treatment. 


**What does `species_nameEschscholzia californica` represent?**

difference between Eschscholzia mass in the well watered and Lotus mass in the well watered


## Stop and think: what does this model mean?  

**What is the best model?**

The best model includes water treatment and species as predictors. 

**How much variation in the response (total mass, in grams) does this model explain?**

72% of the variation in plant mass can be explained by this model.

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

Treatment influences plant mass.
Species influences plant mass.


# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

```{r}
#| label: generate predictions

pred <- ggpredict(
  model4,
  terms = c("water_treatment", "species_name")) |> 
    rename(water_treatment = x,
           species_name = group)


```


# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.

```{r}
#| label: final figure
ggplot(data = drought_exp_clean,
       aes(
         x = water_treatment,
         y = total_g,
         color = water_treatment)) +
  geom_jitter(
    height = 0,
    width = 0.2
  ) +
  geom_pointrange(data = pred,
                  aes(water_treatment,
                      y = predicted,
                      ymin = conf.low,
                      ymax = conf.high)) +
  scale_color_manual(values = c(
    "Drought Stressed" = ds_col,
    "Well Watered" = ww_col
  )) +
  labs(x = "Water treatment",
       y = "Total mass (g)") +
facet_wrap(~ species_name,
           nrow = 1)
  
```

# 6. Writing

## Figure caption








## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**test statistic**


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

#### Goal of analysis


#### Model selection


#### Model diagnostics


### Results









